---
title: "Multi-Biofluid Metabolomics Analysis for Neurodegenerative Diseases"
subtitle: "Biocrates AbsoluteIDQ p400 Kit Data Analysis"
author: "Firalis/Lodiag"
date: "`2024-07-30`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
    self_contained: true
  word_document:
    toc: true
  pdf_document:
    toc: true
always_allow_html: true
editor_options:
  markdown:
    wrap: 72
bibliography: references.bib
---

```{r setup, include=FALSE}
library(knitr)
knitr::opts_chunk$set(warning = FALSE, message = FALSE, 
                      collapse = TRUE, comment = "#>")
library(ADDIA4Metabo)
library(dplyr)
library(readxl)

library(tidyverse)
library(ggplot2)
library(pheatmap)
library(RColorBrewer)
library(corrplot)
library(VennDiagram)
library(FactoMineR)
library(factoextra)
library(limma)
library(ComplexHeatmap)
library(circlize)
library(patchwork)
library(scales)
library(viridis)
library(plotly)
library(ggsci)
library(ggpubr)
library(car)
library(multcomp)
library(mixOmics)
```

# 0. DATA IMPORT AND CLEAN

```{r load_data}
urine <- data.table::fread("metabo_urine.csv") |>
    dplyr::rename(sample_id = "Sample.identification",
                  Disease = Dx) |>
    dplyr::mutate(
        Disease = case_match(
            Disease, "mi AD" ~ "miAD",
            .default = Disease
      )
    )

clinical_data <- readxl::read_excel(
    "../ADDIA4Metabo/data-raw/ADDIA_ST0056_clinical_data_202407.xlsx",
    skip = 1)
serum <- data.table::fread("metabo_serum.csv")  |> 
    dplyr::rename(sample_id = sampleID) |>
    dplyr::left_join(ADDIA4Metabo::raw_data$id_mapping, 
                     by = c(sample_id = "New_ID")) |>
    dplyr::left_join(clinical_data |> 
                         dplyr::select(Name, `APO-E Firalis`),
                     by = c("Original_ID" = "Name")) |>
    dplyr::rename("APOE" = `APO-E Firalis`,
                  Disease = Allgr, Sex = Gender) |>
    dplyr::relocate(APOE, .after = Age) |>
    dplyr::select(-c(Plate, Cohort, Original_ID)) 
```

```{r theme_publication}
theme_publication <- theme_minimal() +
  theme(
    text = element_text(size = 12, family = "Arial"),
    axis.title = element_text(size = 14, face = "bold"),
    axis.text = element_text(size = 11),
    legend.title = element_text(size = 12, face = "bold"),
    legend.text = element_text(size = 11),
    strip.text = element_text(size = 11, face = "bold"),
    plot.title = element_text(size = 16, face = "bold", hjust = 0.5),
    panel.grid.minor = element_blank()
  )

disease_colors <- c("HC" = "#2E8B57", "miAD" = "#FF6347", "DLB" = "#4169E1")
biofluid_colors <- c("Urine" = "#8B4513", "Serum" = "#DC143C")
```

# 1. DATA PREPROCESSING AND QUALITY CONTROL

```{r preprocess_function}
preprocess_biofluid_data <- function(data, biofluid_name) {
  cat(sprintf("Processing %s data...\n", biofluid_name))
  data$Biofluid <- biofluid_name
 
  data$Disease <- case_when(
    data$Disease %in% c("HC", "Control") ~ "HC",
    data$Disease %in% c("miAD", "AD", "Alzheimer") ~ "miAD", 
    data$Disease %in% c("DLB") ~ "DLB",
    TRUE ~ data$Disease
  )
  
  data$Disease <- factor(data$Disease, levels = c("HC", "miAD", "DLB"))
  
  data$Sex <- factor(data$Sex, levels = c("MALE", "FEMALE"))
  
  if(biofluid_name == "Serum") {
    colnames(data) <- gsub("\\.", "(", colnames(data))
    colnames(data) <- gsub("\\($", ")", colnames(data))
    colnames(data) <- gsub("\\(([^)]+)\\(", "(\\1:", colnames(data))
  }
  
  return(data)
}
```


```{r preprocess}
urine_processed <- preprocess_biofluid_data(urine, "Urine")
serum_processed <- preprocess_biofluid_data(serum, "Serum")
```


```{r get_columns}
get_metabolite_columns <- function(data) {
  metadata_cols <- c("V1", "sample_id", "Disease", "Sex", "Age", "APOE", "Biofluid")
  metabolite_cols <- setdiff(colnames(data), metadata_cols)
  return(metabolite_cols)
}

urine_metabolites <- get_metabolite_columns(urine_processed)
serum_metabolites <- get_metabolite_columns(serum_processed)

cat(sprintf("Urine metabolites: %d\n", length(urine_metabolites)))
cat(sprintf("Serum metabolites: %d\n", length(serum_metabolites)))
```

# 2. DESCRIPTIVE STATISTICS AND SAMPLE CHARACTERISTICS

```{r descriptive}
generate_sample_summary <- function(data, biofluid_name) {
  cat(sprintf("\n=== %s Sample Summary ===\n", biofluid_name))
  
  disease_summary <- data %>%
    group_by(Disease) %>%
    summarise(
      n = n(),
      age_mean = mean(Age, na.rm = TRUE),
      age_sd = sd(Age, na.rm = TRUE),
      female_n = sum(Sex == "FEMALE", na.rm = TRUE),
      male_n = sum(Sex == "MALE", na.rm = TRUE),
      .groups = 'drop'
    ) %>%
    mutate(
      female_percent = round(female_n / (female_n + male_n) * 100, 1),
      age_summary = sprintf("%.1f ± %.1f", age_mean, age_sd)
    )
  
  print(disease_summary)
  
  apoe_summary <- data %>%
    group_by(Disease, APOE) %>%
    summarise(n = n(), .groups = 'drop') %>%
    pivot_wider(names_from = APOE, values_from = n, values_fill = 0)
  
  cat("\nAPOE genotype distribution:\n")
  print(apoe_summary)
  
  return(disease_summary)
}

urine_summary <- generate_sample_summary(urine_processed, "Urine")
serum_summary <- generate_sample_summary(serum_processed, "Serum")
```

# 3. METABOLITE CLASSIFICATION AND PATHWAY ANALYSIS

```{r metabolite_class}
classify_metabolites <- function(metabolite_names) {
  classification <- tibble(
    metabolite = metabolite_names,
    class = case_when(
      str_detect(metabolite, "^AC\\(") ~ "Acylcarnitines",
      str_detect(metabolite, "^(Ala|Arg|Asn|Asp|Cit|Gln|Glu|Gly|His|Ile|Lys|Met|Orn|Phe|Pro|Ser|Thr|Trp|Tyr|Val|xLeu)") ~ "Amino Acids",
      str_detect(metabolite, "^(ADMA|SDMA|Carnosine|Creatinine|Dopamine|Histamine|Kynurenine|Met-SO|PEA|Putrescine|Sarcosine|Serotonin|Spermidine|Spermine|t4-OH-Pro|Taurine|Ac-Orn|alpha-AAA|c4-OH-Pro|AcOrn|Nitro-Tyr|t4.OH.Pro|Serotonine)") ~ "Biogenic Amines",
      str_detect(metabolite, "^PC") ~ "Phosphatidylcholines",
      str_detect(metabolite, "^LPC") ~ "Lysophosphatidylcholines", 
      str_detect(metabolite, "^SM") ~ "Sphingomyelins",
      str_detect(metabolite, "^Cer") ~ "Ceramides",
      str_detect(metabolite, "^TG") ~ "Triacylglycerols",
      str_detect(metabolite, "^DG") ~ "Diacylglycerols",
      str_detect(metabolite, "^CE") ~ "Cholesteryl Esters",
      str_detect(metabolite, "Hexose") ~ "Sugars",
      TRUE ~ "Other"
    )
  )
  return(classification)
}

urine_classification <- classify_metabolites(urine_metabolites)
serum_classification <- classify_metabolites(serum_metabolites)

cat("\n=== Metabolite Classification Summary ===\n")
cat("Urine:\n")
print(urine_classification %>% count(class, sort = TRUE))
cat("\nSerum:\n")
print(serum_classification %>% count(class, sort = TRUE))
```

# 4. DATA TRANSFORMATION AND NORMALIZATION

```{r data_transformation}
apply_transformations <- function(data, metabolite_cols) {
  data_transformed <- data
  
  # Log2 transformation for positive values
  for(col in metabolite_cols) {
    values <- data[[col]]
    if(any(values <= 0, na.rm = TRUE)) {
      min_positive <- min(values[values > 0], na.rm = TRUE)
      constant <- min_positive / 2
      values[values <= 0] <- constant
    }
    
    data_transformed[[col]] <- log2(values)
  }
  
  return(data_transformed)
}

urine_transformed <- apply_transformations(urine_processed, urine_metabolites)
serum_transformed <- apply_transformations(serum_processed, serum_metabolites)
```

# 5. UNIVARIATE STATISTICAL ANALYSIS

```{r univariate}
perform_univariate_analysis <- function(data, metabolite_cols, biofluid_name) {
  cat(sprintf("\nPerforming univariate analysis for %s...\n", biofluid_name))
  
  results <- tibble()
  
  for(metabolite in metabolite_cols) {
    values <- data[[metabolite]]
    disease <- data$Disease
    
    complete_idx <- complete.cases(values, disease)
    values_clean <- values[complete_idx]
    disease_clean <- disease[complete_idx]
    
    if(length(unique(disease_clean)) < 2) next
    
    tryCatch({
      aov_result <- aov(values_clean ~ disease_clean)
      aov_summary <- summary(aov_result)
      
      overall_p <- aov_summary[[1]][1, "Pr(>F)"]
      
      ss_treatment <- aov_summary[[1]][1, "Sum Sq"]
      ss_total <- sum(aov_summary[[1]][, "Sum Sq"])
      eta_squared <- ss_treatment / ss_total
      
      tukey_result <- TukeyHSD(aov_result)
      pairwise_p <- tukey_result$disease_clean[, "p adj"]
      
      group_stats <- data %>%
        filter(complete.cases(.[[metabolite]])) %>%
        group_by(Disease) %>%
        summarise(
          mean_val = mean(.data[[metabolite]], na.rm = TRUE),
          sd_val = sd(.data[[metabolite]], na.rm = TRUE),
          n = n(),
          .groups = 'drop'
        )
      
      result_row <- tibble(
        biofluid = biofluid_name,
        metabolite = metabolite,
        overall_p = overall_p,
        eta_squared = eta_squared,
        miAD_vs_HC_p = pairwise_p["miAD-HC"],
        DLB_vs_HC_p = pairwise_p["DLB-HC"], 
        DLB_vs_miAD_p = pairwise_p["DLB-miAD"]
      )
      
      for(disease_group in group_stats$Disease) {
        stats <- group_stats %>% filter(Disease == disease_group)
        result_row[[paste0(disease_group, "_mean")]] <- stats$mean_val
        result_row[[paste0(disease_group, "_sd")]] <- stats$sd_val
        result_row[[paste0(disease_group, "_n")]] <- stats$n
      }
      
      results <- bind_rows(results, result_row)
      
    }, error = function(e) {
      cat(sprintf("Error analyzing %s: %s\n", metabolite, e$message))
    })
  }
  
  if(nrow(results) > 0) {
    results$overall_p_fdr <- p.adjust(results$overall_p, method = "fdr")
    results$miAD_vs_HC_p_fdr <- p.adjust(results$miAD_vs_HC_p, method = "fdr")
    results$DLB_vs_HC_p_fdr <- p.adjust(results$DLB_vs_HC_p, method = "fdr")
    results$DLB_vs_miAD_p_fdr <- p.adjust(results$DLB_vs_miAD_p, method = "fdr")
  }
  
  return(results)
}

urine_univariate <- perform_univariate_analysis(urine_transformed, urine_metabolites, "Urine")
serum_univariate <- perform_univariate_analysis(serum_transformed, serum_metabolites, "Serum")

all_univariate <- bind_rows(urine_univariate, serum_univariate)

significant_metabolites <- all_univariate %>%
  filter(overall_p_fdr < 0.05) %>%
  arrange(overall_p_fdr)

cat(sprintf("\nSignificant metabolites (FDR < 0.05): %d\n", nrow(significant_metabolites)))
print(head(significant_metabolites, 20))
```

# 6. MULTIVARIATE ANALYSIS - PRINCIPAL COMPONENT ANALYSIS

```{r pca}
perform_pca_analysis <- function(data, metabolite_cols, biofluid_name) {
  metabolite_matrix <- data |> dplyr::select(metabolite_cols) 
  
  missing_prop <- apply(metabolite_matrix, 2, function(x) sum(is.na(x))/length(x))
  valid_cols <- names(missing_prop)[missing_prop < 0.3]
  
  metabolite_matrix <- metabolite_matrix |> dplyr::select(valid_cols)
  
  complete_rows <- complete.cases(metabolite_matrix)
  metabolite_matrix <- metabolite_matrix[complete_rows, , drop = FALSE]
  metabolite_matrix <- as.matrix(metabolite_matrix)
  metabolite_matrix <- metabolite_matrix[, apply(metabolite_matrix, 2, function(x) all(is.finite(x))), drop = FALSE]
  metadata <- data[complete_rows, c("Disease", "Sex", "Age", "APOE")]
  
  pca_result <- PCA(metabolite_matrix, graph = FALSE, scale.unit = TRUE)
  
  pca_data <- tibble(
    PC1 = pca_result$ind$coord[, 1],
    PC2 = pca_result$ind$coord[, 2],
    PC3 = pca_result$ind$coord[, 3],
    Disease = metadata$Disease,
    Sex = metadata$Sex,
    Age = metadata$Age,
    APOE = metadata$APOE,
    Biofluid = biofluid_name
  )
  
  return(list(pca_result = pca_result, pca_data = pca_data))
}

urine_pca <- perform_pca_analysis(urine_transformed, urine_metabolites, "Urine")
serum_pca <- perform_pca_analysis(serum_transformed, serum_metabolites, "Serum")
```

# 7. PARTIAL LEAST SQUARES DISCRIMINANT ANALYSIS (PLS-DA)

```{r pls-da}
perform_plsda_analysis <- function(data, metabolite_cols, biofluid_name) {
  metabolite_matrix <- data |> dplyr::select(metabolite_cols)
  
  missing_prop <- apply(metabolite_matrix, 2, function(x) sum(is.na(x))/length(x))
  valid_cols <- names(missing_prop)[missing_prop < 0.3]
  
  metabolite_matrix <- metabolite_matrix |> dplyr::select(valid_cols)
  
  complete_rows <- complete.cases(metabolite_matrix)
  X <- as.matrix(metabolite_matrix[complete_rows, ])
  Y <- data$Disease[complete_rows]
  
  for (i in 1:ncol(X)) {
      if (any(is.na(X[, i]))) {
          X[is.na(X[, i]), i] <- median(X[, i], na.rm = TRUE)
      }
  }
  
  plsda_result <- plsda(X, Y, ncomp = 3)
  
  plsda_data <- tibble(
    Comp1 = plsda_result$variates$X[, 1],
    Comp2 = plsda_result$variates$X[, 2], 
    Comp3 = plsda_result$variates$X[, 3],
    Disease = Y,
    Biofluid = biofluid_name
  )
  
  return(list(plsda_result = plsda_result, plsda_data = plsda_data))
}

urine_plsda <- perform_plsda_analysis(urine_transformed, urine_metabolites, "Urine")
serum_plsda <- perform_plsda_analysis(serum_transformed, serum_metabolites, "Serum")
```

# 8. VISUALIZATION FUNCTIONS

```{r visualization_functions}
create_pca_plot <- function(pca_data, title) {
  ggplot(pca_data, aes(x = PC1, y = PC2, color = Disease, shape = Disease)) +
    geom_point(size = 3, alpha = 0.7) +
    stat_ellipse(aes(color = Disease), level = 0.95, linetype = "dashed") +
    scale_color_manual(values = disease_colors) +
    scale_shape_manual(values = c(16, 17, 15)) +
    labs(
      title = title,
      x = sprintf("PC1 (%.1f%%)", pca_data$variance_pc1 * 100),
      y = sprintf("PC2 (%.1f%%)", pca_data$variance_pc2 * 100),
      color = "Disease Group",
      shape = "Disease Group"
    ) +
    theme_publication +
    theme(legend.position = "bottom")
}

create_plsda_plot <- function(plsda_data, title) {
  ggplot(plsda_data, aes(x = Comp1, y = Comp2, color = Disease, shape = Disease)) +
    geom_point(size = 3, alpha = 0.7) +
    stat_ellipse(aes(color = Disease), level = 0.95, linetype = "dashed") +
    scale_color_manual(values = disease_colors) +
    scale_shape_manual(values = c(16, 17, 15)) +
    labs(
      title = title,
      x = "Component 1",
      y = "Component 2", 
      color = "Disease Group",
      shape = "Disease Group"
    ) +
    theme_publication +
    theme(legend.position = "bottom")
}

create_volcano_plot <- function(univariate_results, biofluid_name, comparison = "miAD_vs_HC") {
  plot_data <- univariate_results %>%
    filter(biofluid == biofluid_name) %>%
    mutate(
      log2_fc = case_when(
        comparison == "miAD_vs_HC" ~ log2(miAD_mean / HC_mean),
        comparison == "DLB_vs_HC" ~ log2(DLB_mean / HC_mean),
        comparison == "DLB_vs_miAD" ~ log2(DLB_mean / miAD_mean)
      ),
      neg_log10_p = case_when(
        comparison == "miAD_vs_HC" ~ -log10(miAD_vs_HC_p_fdr),
        comparison == "DLB_vs_HC" ~ -log10(DLB_vs_HC_p_fdr), 
        comparison == "DLB_vs_miAD" ~ -log10(DLB_vs_miAD_p_fdr)
      ),
      significance = case_when(
        neg_log10_p > -log10(0.05) & abs(log2_fc) > 0.5 ~ "Significant",
        TRUE ~ "Not Significant"
      )
    ) %>%
    filter(!is.na(log2_fc) & !is.na(neg_log10_p) & is.finite(log2_fc) & is.finite(neg_log10_p))
  
  ggplot(plot_data, aes(x = log2_fc, y = neg_log10_p, color = significance)) +
    geom_point(alpha = 0.6, size = 2) +
    geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "red") +
    geom_vline(xintercept = c(-0.5, 0.5), linetype = "dashed", color = "red") +
    scale_color_manual(values = c("Significant" = "red", "Not Significant" = "gray50")) +
    labs(
      title = sprintf("Volcano Plot: %s (%s)", biofluid_name, comparison),
      x = "Log2 Fold Change",
      y = "-Log10 FDR P-value",
      color = "Significance"
    ) +
    theme_publication +
    theme(legend.position = "bottom")
}
```

# 9. GENERATE COMPREHENSIVE VISUALIZATIONS

```{r compr_visualizations}
create_sample_characteristics_plot <- function() {
  combined_summary <- bind_rows(
    urine_summary %>% mutate(Biofluid = "Urine"),
    serum_summary %>% mutate(Biofluid = "Serum")
  )
  
  p1 <- ggplot(combined_summary, aes(x = Disease, y = n, fill = Biofluid)) +
    geom_col(position = "dodge", alpha = 0.8) +
    geom_text(aes(label = n), position = position_dodge(width = 0.9), vjust = -0.5) +
    scale_fill_manual(values = biofluid_colors) +
    labs(
      title = "Sample Size Distribution",
      x = "Disease Group",
      y = "Number of Samples",
      fill = "Biofluid"
    ) +
    theme_publication
  
  combined_age_data <- bind_rows(
    urine_processed %>% select(Disease, Age) %>% mutate(Biofluid = "Urine"),
    serum_processed %>% select(Disease, Age) %>% mutate(Biofluid = "Serum")
  )
  
  p2 <- ggplot(combined_age_data, aes(x = Disease, y = Age, fill = Disease)) +
    geom_boxplot(alpha = 0.7) +
    facet_wrap(~Biofluid) +
    scale_fill_manual(values = disease_colors) +
    labs(
      title = "Age Distribution by Disease Group",
      x = "Disease Group",
      y = "Age (years)",
      fill = "Disease Group"
    ) +
    theme_publication +
    theme(legend.position = "none")
  
  return(list(sample_size = p1, age_dist = p2))
}

create_multivariate_plots <- function() {
  urine_pca$pca_data$variance_pc1 <- urine_pca$pca_result$eig[1, 2] / 100
  urine_pca$pca_data$variance_pc2 <- urine_pca$pca_result$eig[2, 2] / 100
  serum_pca$pca_data$variance_pc1 <- serum_pca$pca_result$eig[1, 2] / 100  
  serum_pca$pca_data$variance_pc2 <- serum_pca$pca_result$eig[2, 2] / 100
  
  p1 <- create_pca_plot(urine_pca$pca_data, "PCA: Urine Metabolome")
  p2 <- create_pca_plot(serum_pca$pca_data, "PCA: Serum Metabolome")
   
  p3 <- create_plsda_plot(urine_plsda$plsda_data, "PLS-DA: Urine Metabolome")
  p4 <- create_plsda_plot(serum_plsda$plsda_data, "PLS-DA: Serum Metabolome")
  
  return(list(pca_urine = p1, pca_serum = p2, plsda_urine = p3, plsda_serum = p4))
}

create_volcano_plots <- function() {
  p1 <- create_volcano_plot(all_univariate, "Urine", "miAD_vs_HC")
  p2 <- create_volcano_plot(all_univariate, "Serum", "miAD_vs_HC") 
  p3 <- create_volcano_plot(all_univariate, "Urine", "DLB_vs_HC")
  p4 <- create_volcano_plot(all_univariate, "Serum", "DLB_vs_HC")
  
  return(list(urine_ad = p1, serum_ad = p2, urine_dlb = p3, serum_dlb = p4))
}
```

# 10. INTEGRATIVE ANALYSIS ACROSS BIOFLUIDS

```{r integrate_across_biofluids}
identify_common_metabolites <- function(urine_results, serum_results, fdr_threshold = 0.05) {
  urine_sig <- urine_results %>%
    filter(overall_p_fdr < fdr_threshold) %>%
    pull(metabolite)
  
  serum_sig <- serum_results %>%
    filter(overall_p_fdr < fdr_threshold) %>%
    pull(metabolite)
  
  common_metabolites <- intersect(urine_sig, serum_sig)
  
  cat(sprintf("Significant metabolites in urine: %d\n", length(urine_sig)))
  cat(sprintf("Significant metabolites in serum: %d\n", length(serum_sig)))
  cat(sprintf("Common significant metabolites: %d\n", length(common_metabolites)))
  
  return(list(
    urine_sig = urine_sig,
    serum_sig = serum_sig,
    common = common_metabolites
  ))
}

common_sig_metabolites <- identify_common_metabolites(urine_univariate, serum_univariate)

perform_cross_biofluid_correlation <- function() {
  if(length(common_sig_metabolites$common) == 0) {
    cat("No common significant metabolites found for correlation analysis.\n")
    return(NULL)
  }
  
  urine_samples <- urine_processed$sample_id
  serum_samples <- serum_processed$sample_id  
  
  urine_class_means <- urine_transformed %>%
    select(Disease, all_of(urine_metabolites)) %>%
    pivot_longer(cols = -Disease, names_to = "metabolite", values_to = "value") %>%
    left_join(urine_classification, by = "metabolite") %>%
    group_by(Disease, class) %>%
    summarise(mean_value = mean(value, na.rm = TRUE), .groups = 'drop') %>%
    pivot_wider(names_from = class, values_from = mean_value, names_prefix = "Urine_")
  
  serum_class_means <- serum_transformed %>%
    select(Disease, all_of(serum_metabolites)) %>%
    pivot_longer(cols = -Disease, names_to = "metabolite", values_to = "value") %>%
    left_join(serum_classification, by = "metabolite") %>%
    group_by(Disease, class) %>%
    summarise(mean_value = mean(value, na.rm = TRUE), .groups = 'drop') %>%
    pivot_wider(names_from = class, values_from = mean_value, names_prefix = "Serum_")
  
  correlation_data <- full_join(urine_class_means, serum_class_means, by = "Disease")
  
  return(correlation_data)
}

cross_biofluid_corr <- perform_cross_biofluid_correlation()
```

# 11. METABOLITE IMPORTANCE AND FEATURE SELECTION

```{r feature_importance}
get_top_discriminative_metabolites <- function(plsda_result, n_top = 20) {
  loadings <- plsda_result$loadings$X[, 1:2]
  
  # Calculate VIP scores (Variable Importance in Projection)
  vip_scores <- sqrt(rowSums(loadings^2 * 
    matrix(rep(plsda_result$prop_expl_var$X[1:2], each = nrow(loadings)), 
           nrow = nrow(loadings)) * ncol(loadings)))
  
  top_metabolites <- tibble(
    metabolite = rownames(loadings),
    vip_score = vip_scores,
    comp1_loading = loadings[, 1],
    comp2_loading = loadings[, 2]
  ) %>%
    arrange(desc(vip_score)) %>%
    head(n_top)
  
  return(top_metabolites)
}

urine_top_metabolites <- get_top_discriminative_metabolites(urine_plsda$plsda_result)
serum_top_metabolites <- get_top_discriminative_metabolites(serum_plsda$plsda_result)
```

# 12. PATHWAY ENRICHMENT ANALYSIS

```{r enrichment}
# pathway enrichment based on metabolite classes
perform_pathway_analysis <- function(significant_metabolites, classification_data, biofluid_name) {
  sig_classification <- classification_data %>%
    filter(metabolite %in% significant_metabolites) %>%
    count(class, sort = TRUE)
  
  total_classification <- classification_data %>%
    count(class, name = "total")
  
  enrichment_results <- sig_classification %>%
    left_join(total_classification, by = "class") %>%
    mutate(
      biofluid = biofluid_name,
      proportion_significant = n / sum(n),
      proportion_total = total / sum(total),
      enrichment_ratio = proportion_significant / proportion_total,
      fold_enrichment = (n / (sum(n) - n)) / (total / (sum(total) - total))
    ) %>%
    arrange(desc(enrichment_ratio))
  
  return(enrichment_results)
}

urine_pathway_enrichment <- perform_pathway_analysis(
  common_sig_metabolites$urine_sig, 
  urine_classification, 
  "Urine"
)

serum_pathway_enrichment <- perform_pathway_analysis(
  common_sig_metabolites$serum_sig,
  serum_classification,
  "Serum"
)
```

# 13. HEATMAP GENERATION

```{r heatmap}
create_metabolite_heatmap <- function(data, metabolites, biofluid_name, top_n = 50) {
  top_metabolites <- metabolites[1:min(top_n, length(metabolites))]
  
  heatmap_data <- data %>%
    select(Disease, all_of(top_metabolites)) %>%
    group_by(Disease) %>%
    summarise(across(everything(), ~ mean(.x, na.rm = TRUE)), .groups = 'drop') %>%
    column_to_rownames("Disease")
  
  # Z-score normalization
  heatmap_matrix <- scale(t(heatmap_data))
  
  metabolite_classes <- urine_classification %>%
    filter(metabolite %in% rownames(heatmap_matrix)) %>%
    column_to_rownames("metabolite")
  
  class_colors <- rainbow(length(unique(metabolite_classes$class)))
  names(class_colors) <- unique(metabolite_classes$class)
  
  annotation_colors <- list(class = class_colors)
  
  pheatmap(
    heatmap_matrix,
    main = sprintf("Top %d Metabolites - %s", nrow(heatmap_matrix), biofluid_name),
    annotation_row = metabolite_classes,
    annotation_colors = annotation_colors,
    cluster_rows = TRUE,
    cluster_cols = FALSE,
    show_rownames = TRUE,
    show_colnames = TRUE,
    scale = "none",
    color = colorRampPalette(c("blue", "white", "red"))(100),
    fontsize = 8,
    fontsize_row = 6
  )
}
```

# 14. STATISTICAL SUMMARY AND EXPORT FUNCTIONS

```{r stats_summary}
create_results_summary <- function() {
  summary_stats <- tibble(
    Metric = c(
      "Urine Samples (n)",
      "Serum Samples (n)",
      "Urine Metabolites (n)",
      "Serum Metabolites (n)",
      "Urine Significant Metabolites (FDR < 0.05)",
      "Serum Significant Metabolites (FDR < 0.05)",
      "Common Significant Metabolites",
      "HC Urine (n)",
      "miAD Urine (n)", 
      "DLB Urine (n)",
      "HC Serum (n)",
      "miAD Serum (n)",
      "DLB Serum (n)"
    ),
    Value = c(
      nrow(urine_processed),
      nrow(serum_processed),
      length(urine_metabolites),
      length(serum_metabolites),
      length(common_sig_metabolites$urine_sig),
      length(common_sig_metabolites$serum_sig),
      length(common_sig_metabolites$common),
      sum(urine_processed$Disease == "HC", na.rm = TRUE),
      sum(urine_processed$Disease == "miAD", na.rm = TRUE),
      sum(urine_processed$Disease == "DLB", na.rm = TRUE),
      sum(serum_processed$Disease == "HC", na.rm = TRUE),
      sum(serum_processed$Disease == "miAD", na.rm = TRUE),
      sum(serum_processed$Disease == "DLB", na.rm = TRUE)
    )
  )
  
  return(summary_stats)
}

final_summary <- create_results_summary()
```

# 15. BIOMARKER PANEL DEVELOPMENT

```{r biomarker_panel}
develop_biomarker_panels <- function(univariate_results, top_metabolites_list, biofluid_name) {
  biomarker_candidates <- univariate_results %>%
    filter(biofluid == biofluid_name & overall_p_fdr < 0.05) %>%
    arrange(overall_p_fdr) %>%
    select(metabolite, overall_p_fdr, eta_squared) %>%
    head(20)
  
  # VIP scores from PLS-DA
  top_metabolites_df <- top_metabolites_list %>%
    select(metabolite, vip_score)
  
  biomarker_panel <- biomarker_candidates %>%
    left_join(top_metabolites_df, by = "metabolite") %>%
    mutate(
      biofluid = biofluid_name,
      combined_score = scale(1 - overall_p_fdr)[, 1] + 
                       scale(eta_squared)[, 1] + 
                       scale(coalesce(vip_score, 0))[, 1]
    ) %>%
    arrange(desc(combined_score)) %>%
    head(10)
  
  return(biomarker_panel)
}

urine_biomarker_panel <- develop_biomarker_panels(all_univariate, urine_top_metabolites, "Urine")
serum_biomarker_panel <- develop_biomarker_panels(all_univariate, serum_top_metabolites, "Serum")
```

# 16. CROSS-VALIDATION AND MODEL PERFORMANCE

```{r biomarker_panel}
# cross-validation for classification
perform_classification_cv <- function(data, metabolite_cols, biofluid_name, k_folds = 5) {
  metabolite_matrix <- data[, metabolite_cols, drop = FALSE]
  
  # here we remove columns with too many missing values
  missing_prop <- apply(metabolite_matrix, 2, function(x) sum(is.na(x))/length(x))
  valid_cols <- names(missing_prop)[missing_prop < 0.3]
  
  metabolite_matrix <- metabolite_matrix[, valid_cols, drop = FALSE]
  
  # we removed also rows with missing values
  complete_rows <- complete.cases(metabolite_matrix)
  X <- as.matrix(metabolite_matrix[complete_rows, ])
  Y <- data$Disease[complete_rows]
  
  # missing values by median imputation
  for(i in 1:ncol(X)) {
    if(any(is.na(X[, i]))) {
      X[is.na(X[, i]), i] <- median(X[, i], na.rm = TRUE)
    }
  }
  
  # k-fold cross-validation
  set.seed(123)
  folds <- sample(rep(1:k_folds, length.out = nrow(X)))
  
  cv_results <- tibble()
  
  for(fold in 1:k_folds) {
    train_idx <- folds != fold
    test_idx <- folds == fold
    
    X_train <- X[train_idx, ]
    Y_train <- Y[train_idx]
    X_test <- X[test_idx, ]
    Y_test <- Y[test_idx]
    
    tryCatch({
      plsda_model <- plsda(X_train, Y_train, ncomp = 2)
      
      predictions <- predict(plsda_model, X_test)
      predicted_classes <- predictions$class$max.dist[, 2]
      
      accuracy <- sum(predicted_classes == Y_test) / length(Y_test)
      
      cv_results <- bind_rows(cv_results, tibble(
        fold = fold,
        biofluid = biofluid_name,
        accuracy = accuracy,
        n_train = length(Y_train),
        n_test = length(Y_test)
      ))
      
    }, error = function(e) {
      cat(sprintf("Error in fold %d: %s\n", fold, e$message))
    })
  }
  
  cv_summary <- cv_results %>%
    summarise(
      biofluid = first(biofluid),
      mean_accuracy = mean(accuracy, na.rm = TRUE),
      sd_accuracy = sd(accuracy, na.rm = TRUE),
      n_folds = n(),
      .groups = 'drop'
    )
  
  return(list(detailed = cv_results, summary = cv_summary))
}

urine_cv <- perform_classification_cv(urine_transformed, urine_metabolites, "Urine")
serum_cv <- perform_classification_cv(serum_transformed, serum_metabolites, "Serum")
```

# 17. GENERATE FINAL VISUALIZATIONS AND REPORTS

```{r final_visual}
sample_plots <- create_sample_characteristics_plot()
multivariate_plots <- create_multivariate_plots()
volcano_plots <- create_volcano_plots()

cat("\n" + "="*80 + "\n")
cat("COMPREHENSIVE MULTI-BIOFLUID METABOLOMICS ANALYSIS SUMMARY\n")
cat("="*80 + "\n")
print(final_summary)

cat("\n=== CROSS-VALIDATION RESULTS ===\n")
print(bind_rows(urine_cv$summary, serum_cv$summary))

cat("\n=== TOP BIOMARKER CANDIDATES ===\n")
cat("Urine:\n")
print(urine_biomarker_panel %>% select(metabolite, overall_p_fdr, eta_squared, vip_score, combined_score))
cat("\nSerum:\n") 
print(serum_biomarker_panel %>% select(metabolite, overall_p_fdr, eta_squared, vip_score, combined_score))

cat("\n=== PATHWAY ENRICHMENT RESULTS ===\n")
cat("Urine:\n")
print(urine_pathway_enrichment)
cat("\nSerum:\n")
print(serum_pathway_enrichment)
```

# 18. EXPORT RESULTS FOR PUBLICATION

```{r final_visual}
export_results <- function() {
  if(!dir.exists("metabolomics_results")) {
    dir.create("metabolomics_results")
  }
  
  # statistical results
  write_csv(all_univariate, "metabolomics_results/univariate_analysis_results.csv")
  write_csv(significant_metabolites, "metabolomics_results/significant_metabolites.csv")
  write_csv(bind_rows(urine_biomarker_panel, serum_biomarker_panel), 
            "metabolomics_results/biomarker_panels.csv")
  write_csv(bind_rows(urine_pathway_enrichment, serum_pathway_enrichment),
            "metabolomics_results/pathway_enrichment.csv")
  write_csv(final_summary, "metabolomics_results/analysis_summary.csv")
  
  # PCA/PLS-DA results
  write_csv(bind_rows(urine_pca$pca_data, serum_pca$pca_data), 
            "metabolomics_results/pca_scores.csv")
  write_csv(bind_rows(urine_plsda$plsda_data, serum_plsda$plsda_data),
            "metabolomics_results/plsda_scores.csv")
  
  # cross-validation results
  write_csv(bind_rows(urine_cv$detailed, serum_cv$detailed),
            "metabolomics_results/cross_validation_results.csv")
  
  cat("Results exported to 'metabolomics_results' directory\n")
}

export_results()
```

# 19. CREATE PUBLICATION-READY FIGURES

```{r pub_ready_figures}
# Figure 1: Study overview and sample characteristics
create_figure1 <- function() {
  fig1 <- (sample_plots$sample_size / sample_plots$age_dist) +
    plot_annotation(
      title = "Figure 1: Study Design and Sample Characteristics",
      subtitle = "Multi-biofluid metabolomics analysis of neurodegenerative diseases",
      tag_levels = 'A'
    )
  
  ggsave("metabolomics_results/Figure1_Sample_Characteristics.pdf", 
         fig1, width = 12, height = 8, dpi = 300)
  ggsave("metabolomics_results/Figure1_Sample_Characteristics.png", 
         fig1, width = 12, height = 8, dpi = 300)
  
  return(fig1)
}

# Figure 2: Multivariate analysis
create_figure2 <- function() {
  fig2 <- (multivariate_plots$pca_urine + multivariate_plots$pca_serum) /
          (multivariate_plots$plsda_urine + multivariate_plots$plsda_serum) +
    plot_annotation(
      title = "Figure 2: Multivariate Analysis of Metabolomic Profiles",
      subtitle = "PCA and PLS-DA reveal biofluid-specific metabolic signatures",
      tag_levels = 'A'
    )
  
  ggsave("metabolomics_results/Figure2_Multivariate_Analysis.pdf",
         fig2, width = 16, height = 12, dpi = 300)
  ggsave("metabolomics_results/Figure2_Multivariate_Analysis.png",
         fig2, width = 16, height = 12, dpi = 300)
  
  return(fig2)
}

# Figure 3: Differential metabolite analysis
create_figure3 <- function() {
  fig3 <- (volcano_plots$urine_ad + volcano_plots$serum_ad) /
          (volcano_plots$urine_dlb + volcano_plots$serum_dlb) +
    plot_annotation(
      title = "Figure 3: Differential Metabolite Analysis",
      subtitle = "Volcano plots showing disease-associated metabolic alterations",
      tag_levels = 'A'
    )
  
  ggsave("metabolomics_results/Figure3_Volcano_Plots.pdf",
         fig3, width = 16, height = 12, dpi = 300)
  ggsave("metabolomics_results/Figure3_Volcano_Plots.png", 
         fig3, width = 16, height = 12, dpi = 300)
  
  return(fig3)
}

figure1 <- create_figure1()
figure2 <- create_figure2()  
figure3 <- create_figure3()
```

# 20. FINAL ANALYSIS SUMMARY AND INTERPRETATION

```{r final_summary}
generate_final_report <- function() {
  report <- paste0(
    "COMPREHENSIVE MULTI-BIOFLUID METABOLOMICS ANALYSIS\n",
    "Biocrates AbsoluteIDQ p400 Kit Analysis\n",
    "Date: ", Sys.Date(), "\n\n",
    
    "STUDY OVERVIEW:\n",
    "- Disease groups: HC (Healthy Controls), miAD (mild Alzheimer's Disease), DLB (Dementia with Lewy Bodies)\n",
    "- Biofluids analyzed: Urine and Serum\n",
    "- Total samples: ", nrow(urine_processed) + nrow(serum_processed), "\n",
    "- Total unique metabolites: ", length(unique(c(urine_metabolites, serum_metabolites))), "\n\n",
    
    "KEY FINDINGS:\n",
    "1. Significant metabolites (FDR < 0.05):\n",
    "   - Urine: ", length(common_sig_metabolites$urine_sig), " metabolites\n",
    "   - Serum: ", length(common_sig_metabolites$serum_sig), " metabolites\n",
    "   - Common across biofluids: ", length(common_sig_metabolites$common), " metabolites\n\n",
    
    "2. Classification performance (cross-validation):\n",
    "   - Urine accuracy: ", sprintf("%.2f ± %.2f", urine_cv$summary$mean_accuracy, urine_cv$summary$sd_accuracy), "\n",
    "   - Serum accuracy: ", sprintf("%.2f ± %.2f", serum_cv$summary$mean_accuracy, serum_cv$summary$sd_accuracy), "\n\n",
    
    "3. Top enriched metabolite classes:\n",
    "   - Urine: ", paste(head(urine_pathway_enrichment$class, 3), collapse = ", "), "\n",
    "   - Serum: ", paste(head(serum_pathway_enrichment$class, 3), collapse = ", "), "\n\n",
    
    "BIOMARKER PANELS:\n",
    "Top 5 urine biomarkers: ", paste(head(urine_biomarker_panel$metabolite, 5), collapse = ", "), "\n",
    "Top 5 serum biomarkers: ", paste(head(serum_biomarker_panel$metabolite, 5), collapse = ", "), "\n\n",
    
    "STATISTICAL METHODS:\n",
    "- Data transformation: Log2 transformation\n",
    "- Univariate analysis: ANOVA with Tukey post-hoc tests\n",
    "- Multiple testing correction: False Discovery Rate (FDR)\n",
    "- Multivariate analysis: PCA and PLS-DA\n",
    "- Cross-validation: 5-fold cross-validation\n",
    "- Effect size: Eta-squared\n",
    "- Variable importance: VIP scores from PLS-DA\n\n",
    
    "FILES GENERATED:\n",
    "- Statistical results: univariate_analysis_results.csv\n",
    "- Significant metabolites: significant_metabolites.csv\n",
    "- Biomarker panels: biomarker_panels.csv\n",
    "- Pathway enrichment: pathway_enrichment.csv\n",
    "- Multivariate scores: pca_scores.csv, plsda_scores.csv\n",
    "- Publication figures: Figure1-3 (PDF and PNG formats)\n"
  )
  
  writeLines(report, "metabolomics_results/Analysis_Report.txt")
  cat(report)
}

generate_final_report()

cat("\n" + "="*80 + "\n")
cat("ANALYSIS COMPLETED SUCCESSFULLY!\n")
cat("All results, figures, and reports have been generated.\n")
cat("Check the 'metabolomics_results' directory for all output files.\n")
cat("="*80 + "\n")
```
